Classifying Protein Secondary Structures through Deep Network Learning

### 2013-12-17 docs/paper.tex
updated paper

\subsection{Verification}
All code for the baseline neural network and the deep network were verified by hand using small examples (no more than sequences of length greater than 5).
Specifically, we traced examples through a perceptron, standard single hidden layer neural network, autoencoder, and stacked autoencoders verifying their correctness.
One detail of note, the small examples that were used for verification consistently required at least 1500 iterations over all the training data in order for the weights to converge.
Therefore, after many hours of debugging, we are confident the functionality of our learners are correct.

@@

\subsection{Dataset}
The datasets used by Rost and Qian were curated with significant domain knowledge therefore we chose to use a combination of these datasets.
Qian's dataset is currently hosted by the University of California - Irvine's Machine Learning Repository \cite{uci}.
We used this data without any modification as it was already divided into a 91 protein training set and a test set of size 15 (resulting in 3500 windows of size 13).
The Rost data consists of 124 proteins.
Unfortunately, the exact data is not hosted for public use.
Therefore, we gathered the listed proteins using the DSSP database \cite{DSSP} to provide labels.
We followed the procedure in \cite{Rost} to group the eight secondary structure classes provided by DSSP into the three standard classes mentioned above.

\begin{figure}[ht!]
\centering
\includegraphics[width=65mm]{.png}
\caption{Informative caption here}
\label{Label here}
\end{figure}

@@

>>>>

\subsection{Verification}
All code for the baseline neural network and the deep network were verified by hand using small examples (no more than sequences of length greater than 5).
Specifically, we traced examples through a perceptron, standard single hidden layer neural network, autoencoder, and stacked autoencoders verifying their correctness.
In these tests, we found that small examples that were used for verification consistently required at least 1500 iterations over all the training data in order for the weights to converge.
Also, in the single-layer neural network test, we found that if the hidden layer size was too small, the autoencoder may be capable of re-creating the input, and each layer may converge, but the representation would be too compact for the full neural
Therefore, after many hours of debugging, we are confident the functionality of our learners are correct.

@@

\subsection{Dataset}
The datasets used by Rost and Qian were curated with significant domain knowledge therefore we chose to use a combination of these datasets.
Qian's dataset is currently hosted by the University of California - Irvine's (UCI) Machine Learning Repository \cite{uci}.
We used this data without any modification as it was already divided into a 91 protein training set and a test set of size 15 (resulting in 3500 windows of size 13).
The Rost data consists of 124 proteins.
Unfortunately, the exact data is not hosted for public use.
Therefore, we gathered the listed proteins using the Define Secondary Structure of Proteins (DSSP) database \cite{DSSP} \cite{DSSP2} to provide labels.
We followed the procedure in \cite{Rost} to group the eight secondary structure classes provided by DSSP into the three standard classes mentioned above.
We combined the Rost dataset with the Qian training data to create our training dataset, and used the UCI test set as our final test set.
Our parameter tuning experiments were tested using internal seven- or five- fold cross validation on the training data set.
(Earlier experiments used seven folds, but we switched to five-fold cross validation later due to time complexity.)
After we chose our parameter values, our final experiments were trained on the entire training set and then tested on the held-aside test set.
All plotted points in the following experiments are the average of the accuracies of each fold of the cross validation.

\subsection{Baseline Window Size}
Window size is important because it determines the number of neighbors that can influence an amino acid's classification.
Figure \ref{fig:baseline_window_size} shows that a window size of 13 or 17 was better than a smaller window.
This is to be expected, because a larger window uses results from sets a larger neighborhood around the central amino acid.
However, increasing the window size sharply increased the number of weights that needed to be trained.
As a compromise between accuraccy and performance, we chose to use a window size of 13 for all other experiments.

\begin{figure}[ht!]
\centering
\includegraphics[width=65mm]{results/baseline/baseline_windowSize.png}
\caption{The effect of window size on accuracy, using a standard neural network. A window size of 13 was chosen for moving forward.}
\label{fig:baseline_window_size}
\end{figure}

\subsection{Baseline Hidden Layer Size}
Hidden layer size determines how much we compact the new representation of the inputs.
Figure \ref{fig:baseline_hidden_layer_size} shows that fewer units in the hidden layer represent the data better.
This shows that because the input data is extremely sparse, an entire window of amino acids can be encoded in very few hidden nodes.
Due to time complexity and physical computing resources, we couldn't test any networks with more than 200 hidden units, and could only test a handful of networks with more than 100 units.
Regardless, the data show that smaller numbers of hidden units per layer provide better prediction accuracy.
Based on these results, in subsequent experiments, we do not use more than 50 hidden units per layer.

\begin{figure}[ht!]
\centering
\includegraphics[width=65mm]{results/baseline/baseline_hiddenLayerSize.png}
\caption{The effect of hidden layer size, using a standard neural network.  Hidden layer sizes of less than 50 will be used moving forward.}
\label{fig:baseline_hidden_layer_size}
\end{figure}

\subsection{Deep Network}
In deep network tests, we used results from the baseline tests to set the window size to 13 and to choose a few good hidden layer sizes.
Figure \ref{fig:deep_num_hidden_layers} justifies our use of multiple hidden layer sizes.
It shows that a larger hidden layer is necessary when there are many hidden layers in the network.
This is consistent with the results of our single-layer neural network verification test, where a too-small hidden layer would strongly affect the network performance.  
In fact, we would expect any issues in early layers to be amplified in later layers, and to be even more obvious in networks with more layers.

The data show that the best accuracy occurs for a six hidden layer network.
Without domain knowledge, it's not clear what underlying physical structure each hidden layer represents, so we are unsure why some numbers of layers work better than others.
However, the network performance consistently increases  until it reaches six layers in size, after which point there is a large drop in accuracy.
We believe this is because six layers are capable of creating the most compact representation of the data the network can achieve, and additional layers simply introduce noise.

\begin{figure}[ht!]
\centering
\includegraphics[width=65mm]{results/deep/deep_numLayers.png}
\caption{The effect of hidden layer size and number of hidden layers, using a deep network. The best average accuracy (52.5\%) was achieved using 6 hidden layers with 26 units per layer.}
\label{fig:deep_num_hidden_layers}
\end{figure}

\subsection{Decay}
In all of the deep network tests until this point, the sizes of all the hidden layers in a network were the same.
We modified our algorithm to use a decay factor, so that each hidden layer was some number of units smaller than the previous layer.
The first hidden for these tests has nearly fifty hidden units, so that the final layer sizes are not too small.

We found that these tests worked signifcantly worse than the standard deep networks.
Those tests achieved accuracies of approximately 52.5\%.
With three hidden layers, an initial hidden layer size of 45 and a decay factor of 25\%, we achieved an average accuracy of \textbf{50.37\%}.
With three hidden layers, an initial hidden layer size of 45 and a decay factor of 33\%, we achieved an average accuracy of \textbf{50.22\%}.

We hypothesize that this is because the network is being compacted too much at each step, to the point of losing information.

@@

% Insert and discuss ROC curves -- should I put ALPHA, BETA, LOOP on one plot?

As we mentioned in verication, if we used too few iterations for the network to converge, we saw a significant decrease in the performance of the network.
